Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-08-17 paper/manuscript.tex
Partially resolve changes to uncertainty estimation by @ppxasjsm and @fjclark

\subsection{How reproducible are alchemical free energy calculations?}
\label{subsec:reproducible}
We restrict our analysis here to repetitions of the same calculation performed with precisely the same force field, as different force fields used to describe the same molecule can lead to wide differences in free energies in some cases.  Even within this restriction,
finite computing resources necessarily limit the generated number of uncorrelated samples of potential energy surfaces, and therefore alchemical free energy calculations only give free energy estimates to within finite precision. An important consideration is how reproducible alchemical free energy calculations are in practice. In simple cases such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package, i.e., with a sample standard deviation under 0.01 kcal/mol~\cite{rizzi2019sampl6}.
For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}. 
There is growing literature describing the theoretical underpinning behind such intrinsic non-reproducibility of MD simulation based predictions. Coveney \textit{et al. }~\cite{coveney2016} have noted that, for a complex dynamical system to reach equilibrium, the dynamics must at least be ``mixing'' which is an established result in ergodic theory. A system which is ``mixing'' is also ``ergodic'' by definition but the reverse is not true. This implies that MD trajectories are chaotic, in other words they are extremely sensitive to their starting conditions and hence one-off simulations can never be reproducible. Thus, it is a necessity to perform ensemble simulations in order to perform ensemble averaging to obtain thermodynamic quantities in a reliable manner. Coveney and Wan recently published a book on the probabilistic foundations of MD and how this approach naturally allows for the treatment of uncertainties that arise in the use of the methodology~\cite{MDbook2025}. In practice, in many cases the simplest and most effective way of performing ensemble simulations is to vary only the initial velocities as these yield reproducible (i.e. statistically robust) free energy predictions. Bhati \textit{et al.} developed the TIES method~\cite{bhati2017} which is based on the use of ensemble methods. It produces excellent results for large scale benchmarking datasets for both RBFEs~\cite{bhati2022} and ABFEs~\cite{bhati2025}. It should be noted that the size of ensemble, that is, the number of independent simulations, is an adjustable parameter whose optimal value is system-dependent and will also be affected by the desired level of precision. Further, this method is extremely flexible and adaptive. For example, one can simply add additional ensemble members as necessary or desired, without having to re-run anything done initially. It should also be noted that ensemble simulations are computationally more effective than long simulations. On a large supercomputer, all the sampling can be done concurrently obtaining the statistical results in the time it takes to run a single simulation. Studies have shown that, in general, an ensemble size of 5 and 10 members is a reasonable starting point for alchemical RBFE and ABFE calculations, respectively, but recommendations are provided to adjust them based on some rules of thumb~\cite{bhati2022, bhati2025}. 
The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand. When practical, a more stringent test is to use different input coordinates for each replica as well as different velocities.

Note that these types of statistical differences concern calculations carried out with a single software package, but simulation package variations can introduce additional discrepancies. Such issues of reproducibility of free energy calculations across different simulation packages have attracted attention recently~\cite{loeffler2018reproducibility,rizzi2019sampl6}. Greater variability is expected between packages due to methodological differences such as integrators, thermostats, barostats, treatment of long-range electrostatics, and potentially other factors. For absolute and relative hydration free energies of small organic molecules a variability of ca. 0.2 kcal/mol between popular simulation packages has been reported~\cite{loeffler2018reproducibility}. In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though the study sought to use identical input and simulation parameters~\cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so. 
A recent study demonstrates reproducibility across software packages as well as free energy estimators, but only when ensemble simulations are employed~\cite{wade2022}. The authors reported results from two different MD engines (NAMD and OpenMM) as well as two free energy estimators (FEP and MBAR) for a set of 54 protein-ligand complexes which were found to be in good agreement when ensemble simulations were used to control uncertainties. To conclude, ensemble simulations can remove the aleatoric uncertainties in MD based predictions, however, further work is needed to ensure reproducibility across software packages arising due to systematic/parametric uncertainties.

\subsection{Is my problem suitable for alchemical free energy calculations?}
\label{subsec:suitability}
Before even planning free energy calculations to study binding to a
particular target, it is important to assess what is known about the
system and its timescales and its suitability for free energy
calculations, as well as the \emph{purpose} of the calculations and
the amount of available computer resources. In some cases, predicting accurate binding free energies for a particular target might be
\emph{more} challenging than simply measuring them! This is
often the case when dealing with database screening problems, where
compounds might be easily and quickly available commercially for
testing and free energy calculations could consume far more resources. Free energy calculations thus typically only
appeal when (slow or costly) synthesis would be required or experiments are otherwise cost-prohibitive.

@@

An additional consideration arises for single topology relative free energy calculations. In this class of alchemical free energy calculations it is necessary to generate a molecular topology that may describe the initial and final states of the perturbation (see Fig.~\ref{fig:fig_topology}). In cases where the end states have high topological similarity and high structural overlap this is relatively straightforward and typically handled by use of MCSS calculations. In situations where the end state topologies differ significantly, or where there is relatively little spatial overlap between the two end states, some user intervention may be necessary to produce a satisfactory input topology.

@@

In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999role, shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}. The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s middle branch depicts how such a dual topology works. Note that the two groups that are switched on or off never see each other. The 'mixed' intermediate atoms (M) are used as well, but they strictly interpolate between physical and dummy atoms, or vice versa.
Bhati \textit{et al.}~\cite{bhati2017, bieniek2021} presented a variant of the single coordinate dual topology approach which is used in the TIES method. Its distinguishing feature is the application of some filters to remove atoms with differences in atom types and partial charges from the structural MCSS identified to ensure chemical similarity between all atom pairs in the final MCSS across both end-points. This is followed by ensuring identical charges for all MCSS atoms by averaging between the two end-points. The resultant change in the sum of atomic charges for all MCSS atoms is uniformly distributed across all atoms in the corresponding alchemical region conserving the overall ligand charge for both end-points. The approach ensures that the final MCSS is both structurally and chemically identical within thresholds, allowing the use of a unique set of MCSS atoms for both ligands and dual topology for their alchemical parts.

@@

In general, there are many ways to calculate an MCSS. One of the most prominent approaches is to estimate the MCSS with a 2D subgraph isomorphism solver(SIS) using the 2D topological stuructre information of molecules\cite{raymond2002maximum}.
A large number of software tools can compute SIS MCSS matches using different cheminformatics packages. Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup}, ProtoCaller~\cite{suruzhon2020protocaller}, SMArt~\cite{petrov2021perturbation}, and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkckombu~\cite{kawabata20143d} are standalone tools. Schr\"{o}dinger's FEP+ planning tool was originally based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.

@@

One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule. In this case, the factorization of the partition function is never possible \cite{Shobana_2000}. It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches. For more on ring opening/closing, see also below. Nevertheless, DC approaches are susceptible to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g. harmonic bonds) are used \cite{Axelsen_1998,Ries_2022}. These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest. The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} avoid this problem.
Additionally, TIES~\cite{bhati2017, bieniek2021} circumvents this issue by design as it does not need any restraints.

@@

\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms cancelling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution\cite{wang2017accurate}.
TIES also overcomes this issue by using dual topology and ensuring that MCSS atoms are structurally and chemically identical~\cite{bhati2017, bieniek2021}.

@@

There are multiple strategies to alleviate such sampling challenges in an alchemical calculation. A straightforward but computationally expensive method is to run longer simulations with multiple replicates to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}. Another approach, known as integration over parts,\cite{jayachandran2006parallelized}, involves performing separate alchemical calculations limited to a series of binding macrostates and  then computing the overall free energy difference using the free energy combination formula~\cite{gallicchio2011recent,hsu2023alchemical}, i.e., computing the negative log-sum-exp of the Boltzmann weights from each macrostate. In the integration over parts approach, restraining potentials can be used to restrict the sampling to specific macrostates,\cite{khuttan2023taming}. However, in practice, restraining potentials are sometimes omitted when the states are known to be separated by large free energy barriers and do not interconvert within the simulation time scale, such as cases like Fig.~\ref{fig:configurational_sampling}\textbf{B}. A generally more effective strategy is to bias relevant configurational collective variables (CVs) during the alchemical simulation. This can be achieved by enhancing sampling within individual alchemical states and swapping conformations between alchemical states (preferably from states separated by large free energy barriers) via replica exchange protocols. More recently, several approaches have been proposed to simultaneously bias both configurational CV and alchemical variables, which can effectively address both scenarios Fig.~\ref{fig:configurational_sampling}\textbf{B} and \textbf{C}. Examples include alchemical metadynamics~\cite{hsu2023alchemical}, $\lambda$-ABF (adaptive biasing force)~\cite{lagardere2024lambda}, and optimized alchemical enhanced sampling (ACES)~\cite{lee2023aces}  and approaches that integrate well-tempered metadynamics~\cite{barducci2008well} with alchemical methods~\cite{zhou2025zooming, khuttan2024make}, to name a few.

@@

The NEQ approaches offer some advantages over the equilibrium sampling based methods. For example, the NEQ protocol allows equilibrium simulations of each end-state to be performed independently without pre-defining a ligand network: the transitions between the ligands can be generated afterwards. This eliminates the redundant equilibrium sampling if one end state is present in multiple perturbations. Equilibrium sampling based approaches would require sampling the central ligand for each edge, while NEQ would sample it once and connect to the other ligands afterwards. The NEQ protocol also provides a simple approach in combining different structures in free energy calculations, e.g. simulations in states at $\lambda=0$ and $\lambda=1$ may be initialized with the independently generated ligand coordinates. Another example would be using true apo and holo protein structures as the end states to initialize simulations for absolute binding free energy calculations. 
Bhati \textit{et al.} recently published two studies directly comparing the equilibrium (EQ) and NEQ approaches at unprecedented scales for both RBFE and ABFE calculations using their ensemble simulation-based TIES method~\cite{wan2023eqvsneq, bhati2025}. Further, they systematically studied the effect of various parameters involved in setting up calculations for both approaches and recommended optimal protocols in each case for industrial applications. They highlighted several limitations of the NEQ approach that should be accounted for while performing such calculations to ensure reliability of predictions.

@@

\subsubsection{How long should I run my simulation for and what information should be saved?}
\label{sec:sim_length_information_kept}
Before launching alchemical free energy calculations it is wise to consider how convergence and completion will be assessed. Different conditions on when to stop alchemical free energy calculations should be determined, and this may require several iterative checks and therefore modifications to the calculation protocol.
One useful metric to use for termination is the expected or desired uncertainty of a desired free energy estimate, though care must be exercised should the uncertainty estimate prove unreliable.
In particular, if the rate of change in the free energy estimate is significant when this condition is met, the simulation may not be locally converged, and more sampling may be necessary to determine a stable free energy estimate which is no longer changing significantly over time. 
A few studies reported systematic variation of simulation lengths using ensemble simulations over an extended range to provide recommendations on appropriate values applicable for most protein-ligand systems. The authors suggest using 4 ns and 10 ns for RBFE~\cite{bhati2017, bhati2022, wan2023eqvsneq} and ABFE~\cite{bhati2025} calculations respectively. These values may be used as a general guideline.

@@

The condition of minimizing the statistical uncertainty of different free energy estimators below a sufficient threshold should be one metric monitored over the simulation. This can be done through the uncertainty estimator built into certain analysis tools such as MBAR, or through more general statistical tools like bootstrap sampling. It should be noted however, that uncertainty estimates have the same limitations as other metrics of convergence, as they are only an uncertainty based on the phase space sampled so far in a simulation, and cannot account for states not sampled, and it is worth considering that they will be an underestimation of the true uncertainty~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}.
A target statistical uncertainty should be chosen at the onset of the simulation to avoid excessively long simulations, or falling into the trap of running until the free energy estimate is "good enough," which is subjective and has no defined criteria. This could be a fixed value such as $0.20$ kcal/mol, or a functional quantity such as "below $0.5$ kcal/mol and $10\%$ of the free energy estimate." The user does not need to monitor this information in real-time and can choose to run simulations for fixed duration (either time or number of samples) and run analysis on the data collected thus far. If more samples are needed, the simulations can be resumed, or, started again in different initial conditions.
Bhati \textit{et al.} recommended such a scheme for large scale RBFE and ABFE calculations using TIES~\cite{bhati2022, bhati2025, wan2023eqvsneq, coveney2022corners}. They suggest a set of optimal parameters (ensemble size, simulation length, number of lambda-states, number and length of NEQ transitions, etc.) to be used initially. Then, only those systems which do not satisfy pre-defined criteria (uncertainties below 0.5 kcal/mol and overlap between forward and reverse work values in case of NEQ calculations) should be further evaluated using additional calculations (in the way suggested) until these criteria are met. Such a scheme optimizes the overall computational efficiency for the full dataset in large scale applications.

@@

The uncertainty in the free energy, for example, can be estimated in multiple ways, e.g. through standard error propagation methods (including MBAR's estimator, which is based on the same principles as standard error propagation), through bootstrap methods, and through ensemble simulations. However, as we have repeatedly pointed out, only ensemble simulations can capture the true uncertainty in MD based predictions. All other methods underestimate the true uncertainty as reported by several authors~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}. 
Independent of how the property is estimated, it is important to remember that results of any free energy analysis are \textit{estimations of the given property}, not the true underlying value of the property itself. 
These estimators are usually consistent estimators, meaning they will converge to the true answer in the limit of sufficient sampling, not necessarily unbiased ones though.
As such, it is a good idea to subject different estimators to the same data to see if they yield either the same estimate (within error and bias), or if they fluctuate wildly. See, for example, the potential of mean force with respect to $\vec{\lambda}$ estimated from a bound simulation of a Tyk2 ligand pair of Wang et al.~\cite{wang2015accurate} for both the MBAR and TI estimators, as seen in Fig.~\ref{fig:pmf}.
This is not a perfect method as some estimators, such as exponential averaging, will converge significantly more slowly, relative to more accurate estimators like MBAR. 
Therefore, it is a good idea to apply the estimators to different fractions of the data to see if the main estimator of free energy you have chosen is stable.

@@

Compared with TI, the MBAR method (Sec.~\ref{subsec:estimators}) discussed above provides uncertainty estimation directly from solving a set of linear equations to compute the variances between all states. 
The number of states and amount of sampling should be optimized to minimize the uncertainty in the MBAR free energy estimate, while balancing other key considerations such as computational expense. 
However, this process based on one-off simulations grossly underestimates the true uncertainty and hence uncertainties for MBAR also must be obtained using ensemble simulations~\cite{bhati2018, wade2022}.

@@

>>>>

\subsection{How reproducible are alchemical free energy calculations?}
\label{subsec:reproducible}
We restrict our analysis here to repetitions of the same calculation performed with precisely the same force field, as different force fields used to describe the same molecule can lead to wide differences in free energies in some cases.  Even within this restriction,
finite computing resources necessarily limit the generated number of uncorrelated samples of potential energy surfaces, and therefore alchemical free energy calculations only give free energy estimates to within finite precision. An important consideration is how reproducible alchemical free energy calculations are in practice. In simple cases, such as absolute hydration free energies of small organic molecules, or relative hydration free energy calculations between structurally similar small organic molecules, it should be possible to obtain highly precise estimates with a given software package, i.e., with a sample standard deviation under 0.01 kcal/mol~\cite{rizzi2019sampl6}.
For more complex use cases such as protein-ligand binding free energies the repeatability is often substantially worse~\cite{rizzi2019sampl6}.

%There is growing literature describing the theoretical underpinning behind such intrinsic non-reproducibility of MD simulation based predictions. Coveney \textit{et al. }~\cite{coveney2016} have noted that, for a complex dynamical system to reach equilibrium, the dynamics must at least be ``mixing'' which is an established result in ergodic theory. A system which is ``mixing'' is also ``ergodic'' by definition but the reverse is not true. This implies that MD trajectories are chaotic, in other words they are extremely sensitive to their starting conditions and hence one-off simulations can never be reproducible. Thus, it is a necessity to perform ensemble simulations in order to perform ensemble averaging to obtain thermodynamic quantities in a reliable manner. Coveney and Wan recently published a book on the probabilistic foundations of MD and how this approach naturally allows for the treatment of uncertainties that arise in the use of the methodology~\cite{MDbook2025}. In practice, in many cases the simplest and most effective way of performing ensemble simulations is to vary only the initial velocities as these yield reproducible (i.e. statistically robust) free energy predictions. Bhati \textit{et al.} developed the TIES method~\cite{bhati2017} which is based on the use of ensemble methods. It produces excellent results for large scale benchmarking datasets for both RBFEs~\cite{bhati2022} and ABFEs~\cite{bhati2025}. It should be noted that the size of ensemble, that is, the number of independent simulations, is an adjustable parameter whose optimal value is system-dependent and will also be affected by the desired level of precision. Further, this method is extremely flexible and adaptive. For example, one can simply add additional ensemble members as necessary or desired, without having to re-run anything done initially. It should also be noted that ensemble simulations are computationally more effective than long simulations. On a large supercomputer, all the sampling can be done concurrently obtaining the statistical results in the time it takes to run a single simulation. Studies have shown that, in general, an ensemble size of 5 and 10 members is a reasonable starting point for alchemical RBFE and ABFE calculations, respectively, but recommendations are provided to adjust them based on some rules of thumb~\cite{bhati2022, bhati2025}. 
The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand. When practical, a more stringent test is to use different input coordinates for each repeat, or at least different starting velocities. Repeats or ensemble-based simulations will allow for a robust error statistic on single free energy calculations and are typically expected in the order of 1 kcal/mol or better.

%To estimate ensemble averages and provide robust error statistics generally three independent repeats of simulations are recommended. In this case a repeat or replica are taken as synonyms. As such other approaches use even larger ensembles of simulations such as in the TIES method~\cite{bhati2017} developed by Bhati et al., where a minimum of 5 to 10 replicas are recommended to make up the ensemble for RBFE and ABFE calculations respectively~\cite{bhati2022, bhati2025}. Typically single free energy differences across multiple replicas are expected to be reproducible within 1 kcal/mol.

Note that these types of statistical differences concern calculations carried out with a single software package, but simulation package variations can introduce additional discrepancies. Such issues of reproducibility of free energy calculations across different simulation packages have attracted attention recently~\cite{loeffler2018reproducibility,rizzi2019sampl6, wade2022}. Greater variability is expected between packages due to methodological differences such as integrators, thermostats, barostats, treatment of long-range electrostatics, and potentially other factors. For absolute and relative hydration free energies of small organic molecules a variability of ca. 0.2 kcal/mol between popular simulation packages has been reported~\cite{loeffler2018reproducibility}. In the recent SAMPL6 SAMPLing challenge a larger variability of 0.3 to 1.0 kcal/mol was noted in the computed absolute binding free energies of host/guest systems even though the study sought to use identical input and simulation parameters~\cite{rizzi2019sampl6} and, in many cases, single-point energies were identical or nearly so. 
% Add something on RBFE calculations across engines.
% I think one of the points here that may be one for the other paper is that we are really just looking at a per edge DDG basis. 
% - https://chemrxiv.org/engage/api-gateway/chemrxiv/assets/orp/resource/item/684b48f03ba0887c33ef29e7/original/modular-and-interoperable-workflows-for-benchmarking-alchemical-binding-free-energy-calculation-methodologies.pdf
% - other references?

%A recent study demonstrates reproducibility across software packages as well as free energy estimators, but only when ensemble simulations are employed~\cite{wade2022}. The authors reported results from two different MD engines (NAMD and OpenMM) as well as two free energy estimators (FEP and MBAR) for a set of 54 protein-ligand complexes which were found to be in good agreement when ensemble simulations were used to control uncertainties. To conclude, ensemble simulations can remove the aleatoric uncertainties in MD based predictions, however, further work is needed to ensure reproducibility across software packages arising due to systematic/parametric uncertainties.

\subsection{Is my problem suitable for alchemical free energy calculations?}
\label{subsec:suitability}
Before even planning free energy calculations to study binding to a
particular target, it is important to assess what is known about the
system and its timescales and its suitability for free energy
calculations, as well as the \emph{purpose} of the calculations and
the amount of available computer resources. In some cases, predicting accurate binding free energies for a particular target might be
\emph{more} challenging than simply measuring them! This is
often the case when dealing with database screening problems, where
compounds might be easily and quickly available commercially for
testing and free energy calculations could consume far more resources. Free energy calculations typically only
appeal when (slow or costly) synthesis would be required or experiments are otherwise cost-prohibitive.

@@

An additional consideration arises for single topology relative free energy calculations. In this class of alchemical free energy calculations, it is necessary to generate a molecular topology that may describe the initial and final states of the perturbation (see Fig.~\ref{fig:fig_topology}). In cases where the end states have high topological similarity and high structural overlap this is relatively straightforward and typically handled by use of MCSS calculations. In situations where the end state topologies differ significantly, or where there is relatively little spatial overlap between the two end states, some user intervention may be necessary to produce a satisfactory input topology.

@@

In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999role, shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}. The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s middle branch depicts how such a dual topology works. Note that the two groups that are switched on or off never see each other. The 'mixed' intermediate atoms (M) are used as well, but they strictly interpolate between physical and dummy atoms, or vice versa.

@@

In general, there are many ways to calculate an MCSS. One of the most prominent approaches is to estimate the MCSS with a 2D subgraph isomorphism solver(SIS) using the 2D topological stuructre information of molecules\cite{raymond2002maximum}.
A large number of software tools can compute SIS MCSS matches using different cheminformatics packages. Some rely on RDKit~\cite{rdkit2019Dec}, such as pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup}, ProtoCaller~\cite{suruzhon2020protocaller}, SMArt~\cite{petrov2021perturbation}, and partially BioSimSpace~\cite{hedges2019biosimspace}, while others such as fkcombu~\cite{kawabata20143d} are standalone tools. Schr\"{o}dinger's FEP+ planning tool was originally based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}.

@@

One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule. In this case, the factorization of the partition function is never possible \cite{Shobana_2000}. It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches. For more on ring opening/closing, see also below. Nevertheless, DC approaches are susceptible to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g. harmonic bonds) are used \cite{Axelsen_1998,Ries_2022}. These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest. The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} avoid this problem.

@@

\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss}\textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms cancelling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings~\cite{liu2015ring}.
Some approaches have attempted to address this using a single coordinate sets and with soft bond approaches ~\cite{clark2019relative,wang2017accurate} or dual coordinate set approach \cite{jespers2019qligfep, ries2022relative} but a general solution is not yet in mainstream use. Still, FEP+ implements one solution\cite{wang2017accurate}.

@@

There are multiple strategies to alleviate such sampling challenges in an alchemical calculation. A straightforward but computationally expensive method is to run longer simulations with multiple replicates to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}.
If using a large number of replicates these are often termed ensemble approaches and an example of such a sampling scheme is the TIES method~\cite{bhati2017} developed by Bhati et al., where a minimum of 5 to 10 replicas are recommended to make up the ensemble for RBFE and ABFE calculations respectively~\cite{bhati2022, bhati2025}.
Another approach, known as integration over parts~\cite{jayachandran2006parallelized}, involves performing separate alchemical calculations limited to a series of binding macrostates and  then computing the overall free energy difference using the free energy combination formula~\cite{gallicchio2011recent,hsu2023alchemical}, i.e., computing the negative log-sum-exp of the Boltzmann weights from each macrostate. In the integration over parts approach, restraining potentials can be used to restrict the sampling to specific macrostates~\cite{khuttan2023taming}. However, in practice, restraining potentials are sometimes omitted when the states are known to be separated by large free energy barriers and do not interconvert within the simulation time scale, such as cases like Fig.~\ref{fig:configurational_sampling}\textbf{B}.
A generally more effective strategy is to bias relevant configurational collective variables (CVs) during the alchemical simulation. This can be achieved by enhancing sampling within individual alchemical states and swapping conformations between alchemical states (preferably from states separated by large free energy barriers) via replica exchange protocols. More recently, several approaches have been proposed to simultaneously bias both configurational CV and alchemical variables, which can effectively address both scenarios Fig.~\ref{fig:configurational_sampling}\textbf{B} and \textbf{C}. Examples include alchemical metadynamics~\cite{hsu2023alchemical}, $\lambda$-ABF (adaptive biasing force)~\cite{lagardere2024lambda}, and optimized alchemical enhanced sampling (ACES)~\cite{lee2023aces}  and approaches that integrate well-tempered metadynamics~\cite{barducci2008well} with alchemical methods~\cite{zhou2025zooming, khuttan2024make}, to name a few.

@@

The NEQ approaches offer some advantages over the equilibrium sampling based methods. For example, the NEQ protocol allows equilibrium simulations of each end-state to be performed independently without pre-defining a ligand network: the transitions between the ligands can be generated afterwards. This eliminates the redundant equilibrium sampling if one end state is present in multiple perturbations. Equilibrium sampling based approaches would require sampling the central ligand for each edge, while NEQ would sample it once and connect to the other ligands afterwards. The NEQ protocol also provides a simple approach in combining different structures in free energy calculations, e.g. simulations in states at $\lambda=0$ and $\lambda=1$ may be initialized with the independently generated ligand coordinates. Another example would be using true apo and holo protein structures as the end states to initialize simulations for absolute binding free energy calculations. 
%Bhati \textit{et al.} recently published two studies directly comparing the equilibrium (EQ) and NEQ approaches at unprecedented scales for both RBFE and ABFE calculations using their ensemble simulation-based TIES method~\cite{wan2023eqvsneq, bhati2025}. Further, they systematically studied the effect of various parameters involved in setting up calculations for both approaches and recommended optimal protocols in each case for industrial applications. They highlighted several limitations of the NEQ approach that should be accounted for while performing such calculations to ensure reliability of predictions.

@@

\subsubsection{How long should I run my simulation for and what information should be saved?}
\label{sec:sim_length_information_kept}
Before launching alchemical free energy calculations it is wise to consider how convergence and completion will be assessed. Different conditions on when to stop alchemical free energy calculations should be determined, and this may require several iterative checks and therefore modifications to the calculation protocol.
One useful metric to use for termination is the expected or desired uncertainty of a desired free energy estimate, though care must be exercised should the uncertainty estimate prove unreliable.
In particular, if the rate of change in the free energy estimate is significant when this condition is met, the simulation may not be locally converged, and more sampling may be necessary to determine a stable free energy estimate which is no longer changing significantly over time. 
%A few studies reported systematic variation of simulation lengths using ensemble simulations over an extended range to provide recommendations on appropriate values applicable for most protein-ligand systems. The authors suggest using 4 ns and 10 ns for RBFE~\cite{bhati2017, bhati2022, wan2023eqvsneq} and ABFE~\cite{bhati2025} calculations respectively. These values may be used as a general guideline.

@@

The condition of minimizing the statistical uncertainty of different free energy estimators below a sufficient threshold should be one metric monitored over the simulation. This can be done through the uncertainty estimator built into certain analysis tools such as MBAR, or through more general statistical tools like bootstrap sampling. It should be noted however, that uncertainty estimates have the same limitations as other metrics of convergence, as they are only an uncertainty based on the phase space sampled so far in a simulation, and cannot account for states not sampled, and it is worth considering that they will be an underestimation of the true uncertainty~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}.
A target statistical uncertainty should be chosen at the onset of the simulation to avoid excessively long simulations, or falling into the trap of running until the free energy estimate is "good enough," which is subjective and has no defined criteria. This could be a fixed value such as $0.20$ kcal/mol, or a functional quantity such as "below $0.5$ kcal/mol and $10\%$ of the free energy estimate." The user does not need to monitor this information in real-time and can choose to run simulations for fixed duration (either time or number of samples) and run analysis on the data collected thus far. If more samples are needed, the simulations can be resumed, or, started again in different initial conditions enlarging the overall simulation ensemble. 
%Bhati \textit{et al.} recommended such a scheme for large scale RBFE and ABFE calculations using TIES~\cite{bhati2022, bhati2025, wan2023eqvsneq, coveney2022corners}. They suggest a set of optimal parameters (ensemble size, simulation length, number of lambda-states, number and length of NEQ transitions, etc.) to be used initially. Then, only those systems which do not satisfy pre-defined criteria (uncertainties below 0.5 kcal/mol and overlap between forward and reverse work values in case of NEQ calculations) should be further evaluated using additional calculations (in the way suggested) until these criteria are met. Such a scheme optimizes the overall computational efficiency for the full dataset in large scale applications.

@@

The uncertainty in the free energy, for example, can be estimated in multiple ways, e.g. through standard error propagation methods (including MBAR's estimator, which is based on the same principles as standard error propagation), through bootstrap methods, and through ensemble simulations. 
%However, as we have repeatedly pointed out, only ensemble simulations can capture the true uncertainty in MD based predictions. All other methods underestimate the true uncertainty as reported by several authors~\cite{wan2020fep+, wan2021uq, vassaux2021, gapsys2020large, wade2022, bhati2017, bhati2022, bhati2025, wan2023eqvsneq, bhati2018, bhati2019}. 
Independent of how the property is estimated, it is important to remember that results of any free energy analysis are \textit{estimations of the given property}, not the true underlying value of the property itself. 
These estimators are usually consistent estimators, meaning they will converge to the true answer in the limit of sufficient sampling, not necessarily unbiased ones though.
As such, it is a good idea to subject different estimators to the same data to see if they yield either the same estimate (within error and bias), or if they fluctuate wildly. See, for example, the potential of mean force with respect to $\vec{\lambda}$ estimated from a bound simulation of a Tyk2 ligand pair of Wang et al.~\cite{wang2015accurate} for both the MBAR and TI estimators, as seen in Fig.~\ref{fig:pmf}.
This is not a perfect method as some estimators, such as exponential averaging, will converge significantly more slowly, relative to more accurate estimators like MBAR. 
Therefore, it is a good idea to apply the estimators to different fractions of the data to see if the main estimator of free energy you have chosen is stable.

@@

Compared with TI, the MBAR method (Sec.~\ref{subsec:estimators}) discussed above provides uncertainty estimation directly from solving a set of linear equations to compute the variances between all states. 
The number of states and amount of sampling should be optimized to minimize the uncertainty in the MBAR free energy estimate, while balancing other key considerations such as computational expense. A less accurate uncertainty estimate may be desirable over a reduced computational expense, increased sampling strategies such as using larger simulation ensembles will provide more accurate uncertainties but will drastically increase the computational cost. 
%However, this process based on one-off simulations grossly underestimates the true uncertainty and hence uncertainties for MBAR also must be obtained using ensemble simulations~\cite{bhati2018, wade2022}.

@@

Uncertainty can also be assessed for a particular perturbation by repeating calculations with slight changes in initial configurations, forcefield parameters, and different random seeds in the MD engine. 
The assessment of variability in free energy calculations due to repeating simulations has been previously reported~\cite{aldeghi2019accurate,paliwal2011benchmark,mey2016blinded,mey2018impact}, and large variance in free energies estimated from simulations with different random seeds should be flagged as issues with convergence.
